---
title: "Food Delivery Orders and Customer Behavior Analysis"
author: "mayank saini"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

# 1. Introduction

Food delivery services have become an important part of the modern food and
restaurant industry. Customers can order food from different restaurants
through online delivery platforms, making it possible to analyze ordering
patterns, food costs, delivery performance and customer satisfaction.

This project analyzes food delivery order data using R. The analysis focuses
on order costs, delivery charges, preparation time, delivery time, customer
ratings, cuisines, restaurants and order status.

Statistical analysis and data visualization techniques are used to identify
patterns and relationships in the food delivery data.

# 2. Objectives

The main objectives of this project are:

- To import and understand the food delivery dataset.
- To examine the structure and characteristics of the data.
- To identify missing and duplicate values.
- To clean and preprocess the dataset.
- To analyze food and delivery costs.
- To analyze preparation and delivery times.
- To compare different cuisine types.
- To analyze restaurant performance.
- To study customer ratings.
- To analyze order status.
- To examine relationships between numerical variables.
- To create meaningful graphs and visualizations.
- To identify important findings and provide recommendations.

# 3. Data Importing

## Explanation of the R Code

The `read.csv()` function is used to import the food delivery CSV file into R.
The imported dataset is stored in an object called `food_data`.

```{r data-import, message=FALSE, warning=FALSE}

# Load required package
library(ggplot2)

# Import the CSV dataset
food_data <- read.csv("food_delivery_orders.csv")

# Display the first six records
head(food_data)

```

## Interpretation

The dataset has been successfully imported into R. The first six records are
displayed to understand the type of information available in the dataset.

# 4. Understanding the Dataset

## 4.1 Number of Records

```{r number-records}

cat("Number of records:", nrow(food_data))

```

## 4.2 Number of Variables

```{r number-variables}

cat("Number of variables:", ncol(food_data))

```

## 4.3 Column Names

```{r column-names}

names(food_data)

```

## 4.4 Dataset Structure

```{r dataset-structure}

str(food_data)

```

## 4.5 Summary of Dataset

```{r dataset-summary}

summary(food_data)

```

## Explanation of the R Code

- `nrow()` counts the number of rows in the dataset.
- `ncol()` counts the number of columns.
- `names()` displays the variable names.
- `str()` shows the structure and data type of each variable.
- `summary()` provides descriptive information about the variables.

## Interpretation

The dataset contains information about food delivery orders, including
customer details, restaurants, cuisine types, costs, preparation time,
delivery time, ratings and order status.

# 5. Data Preprocessing and Cleaning

Data cleaning is an important step before statistical analysis. It helps
ensure that the results are based on valid and consistent observations.

## 5.1 Check Missing Values

```{r missing-values}

missing_values <- colSums(is.na(food_data))

missing_values

```

## Interpretation

The number of missing values is displayed for every variable. Variables with
missing values can affect statistical calculations and therefore need to be
considered during preprocessing.

## 5.2 Check Duplicate Records

```{r duplicate-records}

duplicate_count <- sum(duplicated(food_data))

cat("Number of duplicate records:", duplicate_count)

```

## Interpretation

Duplicate records can result in the same order being counted more than once.
The result above shows whether duplicate observations are present.

## 5.3 Check Data Types

```{r data-types}

sapply(food_data, class)

```

## 5.4 Check Unique Order Statuses

```{r order-status-values}

unique(food_data$Order_Status)

```

## 5.5 Check Unique Cuisine Types

```{r cuisine-values}

unique(food_data$Cuisine_Type)

```

# 6. Creating a Clean Dataset

```{r clean-data}

food_clean <- food_data

# Remove duplicate records
food_clean <- food_clean[!duplicated(food_clean), ]

# Remove rows with missing important numerical values
food_clean <- food_clean[
  !is.na(food_clean$Food_Cost) &
  !is.na(food_clean$Delivery_Fee) &
  !is.na(food_clean$Total_Order_Cost) &
  !is.na(food_clean$Prep_Time_Min) &
  !is.na(food_clean$Delivery_Time_Min) &
  !is.na(food_clean$Order_Rating),
]

# Keep positive cost and time values
food_clean <- food_clean[
  food_clean$Food_Cost >= 0 &
  food_clean$Delivery_Fee >= 0 &
  food_clean$Total_Order_Cost >= 0 &
  food_clean$Prep_Time_Min >= 0 &
  food_clean$Delivery_Time_Min >= 0,
]

cat("Original number of records:", nrow(food_data), "\n")
cat("Records after cleaning:", nrow(food_clean), "\n")
cat("Records removed:", nrow(food_data) - nrow(food_clean))

```

## Explanation

The dataset is copied into `food_clean`. Duplicate observations are removed,
followed by observations with missing important numerical values.

Negative values for costs and delivery times are also excluded because they
are not meaningful for this analysis.

## Interpretation

The cleaned dataset contains valid observations that can be used for the
remaining statistical analysis.

# 7. Basic Statistical Analysis

## 7.1 Food Cost Statistics

```{r food-cost-statistics}

food_cost_stats <- c(
  Mean = mean(food_clean$Food_Cost),
  Median = median(food_clean$Food_Cost),
  Minimum = min(food_clean$Food_Cost),
  Maximum = max(food_clean$Food_Cost),
  Standard_Deviation = sd(food_clean$Food_Cost)
)

food_cost_stats

```

## 7.2 Delivery Fee Statistics

```{r delivery-fee-statistics}

delivery_fee_stats <- c(
  Mean = mean(food_clean$Delivery_Fee),
  Median = median(food_clean$Delivery_Fee),
  Minimum = min(food_clean$Delivery_Fee),
  Maximum = max(food_clean$Delivery_Fee),
  Standard_Deviation = sd(food_clean$Delivery_Fee)
)

delivery_fee_stats

```

## 7.3 Total Order Cost Statistics

```{r order-cost-statistics}

order_cost_stats <- c(
  Mean = mean(food_clean$Total_Order_Cost),
  Median = median(food_clean$Total_Order_Cost),
  Minimum = min(food_clean$Total_Order_Cost),
  Maximum = max(food_clean$Total_Order_Cost),
  Standard_Deviation = sd(food_clean$Total_Order_Cost)
)

order_cost_stats

```

## Explanation

The `mean()` function calculates the average, `median()` calculates the
middle value, `min()` and `max()` find the minimum and maximum values, and
`sd()` calculates standard deviation.

## Interpretation

These statistics provide an overview of food costs, delivery charges and
overall order costs in the dataset.

# 8. Order Status Analysis

```{r order-status-table}

status_count <- table(food_clean$Order_Status)

status_count

```

## Order Status Visualization

```{r order-status-graph, fig.width=8, fig.height=5}

status_data <- as.data.frame(status_count)

names(status_data) <- c("Order_Status", "Number_of_Orders")

ggplot(
  status_data,
  aes(x = Order_Status, y = Number_of_Orders)
) +
  geom_col() +
  labs(
    title = "Distribution of Food Delivery Order Status",
    x = "Order Status",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The bar chart shows the number of orders belonging to each order status.
This helps identify the proportion of orders that were successfully delivered
and those that may have experienced another status.

# 9. Cuisine Type Analysis

```{r cuisine-count}

cuisine_count <- table(food_clean$Cuisine_Type)

cuisine_count

```

## Cuisine Visualization

```{r cuisine-graph, fig.width=9, fig.height=6}

cuisine_data <- as.data.frame(cuisine_count)

names(cuisine_data) <- c("Cuisine_Type", "Number_of_Orders")

ggplot(
  cuisine_data,
  aes(
    x = reorder(Cuisine_Type, Number_of_Orders),
    y = Number_of_Orders
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of Orders by Cuisine Type",
    x = "Cuisine Type",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The graph compares the popularity of different cuisine types based on the
number of orders. Cuisine types with larger bars received more orders in
the dataset.

# 10. Restaurant Analysis

## 10.1 Orders by Restaurant

```{r restaurant-count}

restaurant_count <- table(food_clean$Restaurant_Name)

restaurant_data <- as.data.frame(restaurant_count)

names(restaurant_data) <- c(
  "Restaurant_Name",
  "Number_of_Orders"
)

restaurant_data <- restaurant_data[
  order(-restaurant_data$Number_of_Orders),
]

head(restaurant_data, 10)

```

## 10.2 Top Restaurants

```{r restaurant-graph, fig.width=9, fig.height=6}

top_restaurants <- head(restaurant_data, 10)

ggplot(
  top_restaurants,
  aes(
    x = reorder(Restaurant_Name, Number_of_Orders),
    y = Number_of_Orders
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Restaurants by Number of Orders",
    x = "Restaurant",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The graph identifies the restaurants with the highest number of orders.
This provides an indication of restaurant popularity within the dataset.

# 11. Food Cost Analysis

```{r food-cost-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(x = Food_Cost)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Food Costs",
    x = "Food Cost",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The histogram shows how food costs are distributed among the orders.
It helps identify whether most orders are concentrated within a particular
cost range.

# 12. Total Order Cost Analysis

```{r total-cost-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(x = Total_Order_Cost)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Total Order Cost",
    x = "Total Order Cost",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The histogram displays the distribution of total spending per order.
It provides an overview of customer spending patterns.

# 13. Delivery Time Analysis

## 13.1 Delivery Time Statistics

```{r delivery-time-statistics}

delivery_time_stats <- c(
  Mean = mean(food_clean$Delivery_Time_Min),
  Median = median(food_clean$Delivery_Time_Min),
  Minimum = min(food_clean$Delivery_Time_Min),
  Maximum = max(food_clean$Delivery_Time_Min),
  Standard_Deviation = sd(food_clean$Delivery_Time_Min)
)

delivery_time_stats

```

## 13.2 Delivery Time Distribution

```{r delivery-time-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(x = Delivery_Time_Min)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Delivery Time",
    x = "Delivery Time (Minutes)",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The delivery-time graph shows how long orders generally take to reach
customers. It can help identify common delivery-time ranges and unusually
long deliveries.

# 14. Food Preparation Time Analysis

```{r preparation-time-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(x = Prep_Time_Min)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Food Preparation Time",
    x = "Preparation Time (Minutes)",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The graph shows the distribution of food preparation times. Preparation time
is an important factor that can contribute to the total time required to
complete a food delivery order.

# 15. Customer Rating Analysis

## 15.1 Rating Distribution

```{r rating-table}

rating_count <- table(food_clean$Order_Rating)

rating_count

```

```{r rating-graph, fig.width=8, fig.height=5}

rating_data <- as.data.frame(rating_count)

names(rating_data) <- c(
  "Order_Rating",
  "Number_of_Orders"
)

ggplot(
  rating_data,
  aes(
    x = factor(Order_Rating),
    y = Number_of_Orders
  )
) +
  geom_col() +
  labs(
    title = "Distribution of Customer Ratings",
    x = "Order Rating",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## 15.2 Average Rating

```{r average-rating}

average_rating <- mean(
  food_clean$Order_Rating,
  na.rm = TRUE
)

cat("Average customer rating:", round(average_rating, 2))

```

## Interpretation

The rating distribution shows how customers evaluated their food delivery
orders. The average rating provides an overall measure of customer
satisfaction.

# 16. Customer Analysis

## Number of Unique Customers

```{r customer-count}

number_of_customers <- length(
  unique(food_clean$Customer_ID)
)

cat("Number of unique customers:", number_of_customers)

```

## Orders per Customer

```{r customer-orders}

customer_orders <- table(food_clean$Customer_ID)

customer_orders <- sort(
  customer_orders,
  decreasing = TRUE
)

head(customer_orders, 10)

```

## Interpretation

The customer analysis identifies the number of unique customers and shows
which customers placed the greatest number of orders in the dataset.

# 17. Total Order Cost by Cuisine

```{r cuisine-cost}

cuisine_cost <- aggregate(
  Total_Order_Cost ~ Cuisine_Type,
  data = food_clean,
  FUN = mean
)

names(cuisine_cost)[2] <- "Average_Order_Cost"

cuisine_cost <- cuisine_cost[
  order(-cuisine_cost$Average_Order_Cost),
]

cuisine_cost

```

```{r cuisine-cost-graph, fig.width=9, fig.height=6}

ggplot(
  cuisine_cost,
  aes(
    x = reorder(Cuisine_Type, Average_Order_Cost),
    y = Average_Order_Cost
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Order Cost by Cuisine Type",
    x = "Cuisine Type",
    y = "Average Order Cost"
  ) +
  theme_minimal()

```

## Interpretation

The graph compares average spending across different cuisine types.
This can help identify cuisines associated with higher or lower average
order values.

# 18. Delivery Time and Customer Rating

```{r delivery-rating-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(
    x = Delivery_Time_Min,
    y = Order_Rating
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Relationship Between Delivery Time and Customer Rating",
    x = "Delivery Time (Minutes)",
    y = "Customer Rating"
  ) +
  theme_minimal()

```

## Interpretation

The scatter plot examines whether delivery time is associated with customer
ratings. The trend line provides a simple visual indication of the direction
of the relationship.

A relationship in the graph should not automatically be interpreted as
causation.

# 19. Food Cost and Total Order Cost

```{r cost-relationship, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(
    x = Food_Cost,
    y = Total_Order_Cost
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Relationship Between Food Cost and Total Order Cost",
    x = "Food Cost",
    y = "Total Order Cost"
  ) +
  theme_minimal()

```

## Interpretation

The graph examines the relationship between the food cost and the final
order cost. Since food cost is an important component of the total order
cost, a positive relationship may be expected.

# 20. Correlation Analysis

Correlation measures the strength and direction of a linear relationship
between numerical variables.

```{r correlation-analysis}

correlation_data <- food_clean[
  ,
  c(
    "Food_Cost",
    "Delivery_Fee",
    "Total_Order_Cost",
    "Prep_Time_Min",
    "Delivery_Time_Min",
    "Order_Rating"
  )
]

correlation_matrix <- cor(
  correlation_data,
  use = "complete.obs"
)

round(correlation_matrix, 3)

```

## Explanation

The `cor()` function calculates correlation coefficients.

Correlation values range from -1 to +1:

- A value close to +1 indicates a strong positive relationship.
- A value close to -1 indicates a strong negative relationship.
- A value close to 0 indicates a weak linear relationship.

## Interpretation

The correlation matrix helps identify relationships among food cost,
delivery fee, total order cost, preparation time, delivery time and
customer ratings.

# 21. Total Delivery Time Analysis

The total delivery time represents the complete time involved in preparation
and delivery.

```{r total-delivery-time}

total_delivery_stats <- c(
  Mean = mean(food_clean$Total_Delivery_Time),
  Median = median(food_clean$Total_Delivery_Time),
  Minimum = min(food_clean$Total_Delivery_Time),
  Maximum = max(food_clean$Total_Delivery_Time),
  Standard_Deviation = sd(food_clean$Total_Delivery_Time)
)

total_delivery_stats

```

```{r total-delivery-graph, fig.width=9, fig.height=6}

ggplot(
  food_clean,
  aes(x = Total_Delivery_Time)
) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Total Delivery Time",
    x = "Total Delivery Time (Minutes)",
    y = "Number of Orders"
  ) +
  theme_minimal()

```

## Interpretation

The graph shows the distribution of total time required to prepare and
deliver orders. This is useful for understanding overall delivery efficiency.

# 22. Overall Project Summary

```{r overall-summary}

overall_summary <- data.frame(
  Total_Orders = nrow(food_clean),
  Unique_Customers = length(unique(food_clean$Customer_ID)),
  Unique_Restaurants = length(unique(food_clean$Restaurant_Name)),
  Unique_Cuisines = length(unique(food_clean$Cuisine_Type)),
  Average_Food_Cost = mean(food_clean$Food_Cost),
  Average_Delivery_Fee = mean(food_clean$Delivery_Fee),
  Average_Order_Cost = mean(food_clean$Total_Order_Cost),
  Average_Delivery_Time = mean(food_clean$Delivery_Time_Min),
  Average_Rating = mean(food_clean$Order_Rating)
)

overall_summary

```

# 23. Findings

Based on the statistical analysis and visualizations, the following findings
were identified:

- The dataset contains detailed information about food delivery orders.
- Food costs and total order costs vary across different orders.
- Different cuisine types have different levels of popularity.
- Restaurant order counts vary, indicating differences in customer demand.
- Delivery times vary among orders.
- Food preparation time contributes to the overall delivery process.
- Customer ratings provide an indication of customer satisfaction.
- The relationship between delivery time and customer rating can be examined
  using correlation and scatter-plot analysis.
- Food cost has an important relationship with total order cost.
- The analysis demonstrates how statistical methods and visualization can be
  used to understand food delivery operations and customer behavior.

# 24. Business Recommendations

Based on the analysis, the following recommendations can be considered:

- Restaurants should monitor preparation times to improve delivery efficiency.
- Delivery services should identify orders with unusually long delivery times.
- Restaurants can study popular cuisine categories to understand customer
  demand.
- Customer ratings can be used to identify areas requiring service
  improvement.
- Restaurants can analyze average order values to understand customer
  spending behavior.
- Delivery charges should be monitored to maintain a balance between
  customer satisfaction and operational costs.

# 25. Conclusion

This project analyzed food delivery order data using R.

The analysis included data importing, preprocessing, data cleaning,
descriptive statistics, order analysis, cuisine analysis, restaurant
analysis, cost analysis, delivery-time analysis, customer rating analysis,
customer analysis and correlation analysis.

Various graphs were created to visualize important patterns and relationships
within the dataset. Statistical measures were used to summarize the data and
support the interpretation of results.

Overall, the project demonstrates how R can be used for practical data
analysis and visualization in the food delivery industry. The results can
help understand customer behavior, restaurant demand, order costs and
delivery performance.

# 26. Explanation of R Functions Used

| R Function | Purpose |
|---|---|
| `read.csv()` | Imports CSV data |
| `head()` | Displays first records |
| `nrow()` | Counts rows |
| `ncol()` | Counts columns |
| `names()` | Displays column names |
| `str()` | Shows dataset structure |
| `summary()` | Provides summary statistics |
| `is.na()` | Identifies missing values |
| `duplicated()` | Identifies duplicate records |
| `unique()` | Finds unique values |
| `table()` | Creates frequency tables |
| `mean()` | Calculates average |
| `median()` | Calculates median |
| `min()` | Finds minimum |
| `max()` | Finds maximum |
| `sd()` | Calculates standard deviation |
| `aggregate()` | Calculates grouped statistics |
| `cor()` | Calculates correlation |
| `ggplot()` | Creates visualizations |
| `geom_col()` | Creates bar charts |
| `geom_histogram()` | Creates histograms |
| `geom_point()` | Creates scatter plots |
| `geom_smooth()` | Adds a trend line |
| `theme_minimal()` | Applies a clean graph theme |

# 27. References

1. R Core Team. R: A Language and Environment for Statistical Computing.
2. Posit. RStudio IDE Documentation.
3. Food Delivery Orders dataset used for this project.